---
title: "Continuous Time Multi-State Models and Microsimulation"
author: Devin Incerti
output: 
  beamer_presentation:
    keep_tex: FALSE
    toc: FALSE
    includes:
      in_header: template.tex
    template: default.tex
    latex_engine: xelatex
---

```{r echo=FALSE}
library("knitr")
read_chunk('multistate.R')
```

## Overview
1. What are multi-state models?

2. How can multi-state models be estimated?

3. How can an estimated model be used to derive quantities of interest for health technology evaluation?

## What is a Multi-State Model?
* Model of how individuals move through multiple states in continuous time
* 3 types of states
    + Initial state
    + Intermediate state
    + Absorbing state
* *Competing risk* models are a special case with one initial state and several mutually exclusive absorbing states
* *Survival analysis* is a special case with one initial state (alive) and one absorbing state (dead)

## Canonical Example: The (Reversible) Illness-Death Model
\centering
\begin{tikzpicture}[
roundnode/.style={circle, draw=green!60, fill=green!5, very thick, minimum size=7mm},
squarednode/.style={rectangle, draw=red!60, fill=red!5, very thick, minimum size=5mm},
align=center,node distance=2.5cm]
%Nodes
\node[squarednode]      (healthy)                       {Inititial State \\(Healthy)};
\node[squarednode]      (sick)       [right=of healthy] {Intermediate State \\(Sick)};
\coordinate             (middle) at ($(healthy)!0.5!(sick)$);
\node[squarednode]      (dead)       [below=of middle] {Absorbing State \\(Dead)};
 
%Lines
\draw[->] (healthy.south) -- ([xshift=-1ex]dead.north); 
\draw[->] (sick.south) -- ([xshift=1ex]dead.north); 
\draw[->] ([yshift=1ex]healthy.east) -- ([yshift=1ex]sick.west);
\draw[->] ([yshift=-1ex]sick.west) -- ([yshift=-1ex]healthy.east);
\end{tikzpicture}

## Advantages for Parameter Estimation
* Properly accounts for competing risks
    + Individuals experiencing an event are no longer "at risk", which
    decreases transition intensities
* Flexible modeling of hazard rates
    + **Time homogeneous markov models**: transition intensities are constant over time
    + **Time inhomogeneous (clock-forward) markov models**: transition intensities depends on         time since entering the \underline{initial} state
    + **Semi-markov (clock-reset) models**: transition intensities depends on time since                entering the \underline{current} state

## Advantages for Health-Economic Modeling
* Model is in continuous time
    + Predicts exact length of time in states
    + Do not need to worry about multiple events occurring during intervals
* Model is at the individual level
    + Can account for treatment heterogeneity
    + Transition intensities can depend on prior history $\rightarrow$ fewer states than in cohort models
    
## Software for Estimating Multi-State Models
* Nearly all of it is written in R
* Prominent R packages
    + *mstate*: non-parametric and semi-parametric models
    + *flexsurv*: parametric models
    + *msm*: missing information (e.g. do not know exact time of each transition)

## A Liver Cirrhosis Example
* To illustrate methods, lets consider a clinical trial where patients received "Placebo"" or "Prednisone"
* Reversible illness death model with 3 states
    + State 1: Normal prothrombin levels
    + State 2: Low prothrombin levels
    + State 3: Death
* 4 possible transitions
    + Normal to Low
    + Low to Normal
    + Low to Death
    + Normal to Death

## Examining the Data
* A patient treated with Prednisone

```{r setup, echo = FALSE, message=F, warning=F, include=FALSE}
```

```{r create_data, echo = FALSE}
```

```{r show_data, echo = FALSE, comment=NA}
```

* Summarizing the transitions
```{r trans, comment=NA, echo=FALSE}
```

## Fitting Multi-State Models
* Joint model if constraints in parameters across transitions; otherwise 
transition specific models are more computationally efficient
* Individuals entering "competing" states treated as censored
* Types of models
    * *Non-parametric*: fit separate Cox model to treatment and control
    * *Semi-parametric*: clock-forward or clock-reset Cox models
    * *Parametric*: clock-forward or clock-reset model with standard survival 
    distributions (e.g. Weibull, Gompertz)

```{r fit_cox, echo = FALSE}
```

```{r fit_weibull, echo = FALSE}
```

```{r fit_gompertz, echo = FALSE}
```

## Why do we Need Parametric Models?
* We have Kaplan-Meier curves and Cox models so why use parametric models?
* Some advantages of parametric approaches
    + Extrapolation beyond time periods in data
    + Prediction for new individuals
    + Less prone to overfitting
    + Faster simulations
    + Time dependent effects (e.g. non-proportional hazards)
* Estimates from flexible parametric models are very similar to non-parametric and semi-parametric alternatives
* Model checking is essential though!

## Some Model Checking
```{r cumhaz_cox, echo = FALSE}
```

```{r cumhaz_parametric, echo = FALSE}
```

```{r cumhaz_plot, echo = FALSE}
```

## Transition Probabilities in Multi-State Models
* A transition probability is the probability of being in state $X(T)=s$ at time $T$ given that an individual is in state $r$ at time $t_0$
\begin{align} \label{eqn:tp}
P_{rs}(t_0, T) = P(X(T)=s|X(t_0) = r)
\end{align}
* Given discount rate, $\beta(t)$, and quality of life weight, $q_s(t)$, discounted QALYs in state $s$ given being in state $r$ at time $t_0$ are
\begin{align} \label{eqn:tp}
\int_{t_0}^{T} q_s(t) \beta(t) P_{rs}(t_0, t)dt
\end{align}
* Total QALYs are calculated by adding up QALYs in each non-absorbing state

## Estimating Transition Probabilities
* Illness-death model example:
    + Probability of transitioning from state 2 at time $t_0$ to state 3 at time $T$ is given by       multiplying:
        1. Probability of transitioning to state 3 at some time before $T$
        2. Probability of remaining in state 2 until making the transition to 3
* General case for Markov models
    + Transition probabilities can be estimated by solving the *Kolmogorov forward equation*
    + Solution can be approximated by the *Aelen-Johansen estimator*
* Simulation needed for Semi-Markov Models
    + Aelen-Johansen estimator does not apply because transition probabilities from intermediate       states depend on when the state was entered

## Transition Probabilities in Clock-Forward Cox Models
Using R function `probtrans` from *mstate* package
```{r cf_cox_tp, echo = FALSE}
```

## Transition Probabilities in Clock-Reset Cox Models
Using R function `mssample` from *mstate* package
```{r cr_cox_tp, echo = FALSE, cache = T}
```

## Lifetime Simulation 
* Parametric models can be used to simulate outcomes beyond time in data 
* I've created function `simMS` for parametric Semi-Markov models
    + Time (and age) move forward as individuals jump from one event to the next
    + Simulates outcomes for a population of individuals with differing covariate values
    + Simulation stops when all individuals reach absorbing states or time has reached                   preset ending time
    + Can be used with any distribution by supplying a function to generate random numbers and         using the custom distribution framework in *flexsurv*
    + Allows analyst to specify different distributions for each transition
    + Only valid for semi-Markov models; use `probtrans` for clock-forward Markov models

## Benefits of Simulation
* Can calculate standard quantities of interest
    + Discounted QALYs (dQALYs)
    + Length of stay (LOS) in each state
    
\begin{center}
```{r cr_parametric_sim,results='asis', echo = FALSE, comment=NA, cache =T}
```    
\end{center}

* Don't need analytical expressions to estimate other more complicated quantities
* Assess uncertainty (e.g. probabilistic sensitivity analysis)
```{r psa, echo = FALSE, comment=NA, cache=T}
```    

## When Can Multi-State Models be Used?
* Requirements
    + Individual time to event data must be available for each transition
* Scenarios in which it is most useful
    + Individuals spend a lot of time in each state
    + Risks of new events depend on time in state 
    + Chance of multiple events within an interval
    + Need estiamtes for representative patient populations
* Some examples
    + *Transplantation*: functioning transplant $\Leftrightarrow$ failed transplant $\Rightarrow$ death
    + *Oncology*: surgery $\Rightarrow$ local recurrence, distant metastasis $\Rightarrow$ death
